Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-08-10 volunteer-clouds-ieee-ci-si.tex
Revised state of the art
Eliminated comments and done almost-final review. This closes #11.

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense than an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; % [Mario] Aqui empieza
for instance numerous vulnerabilities have been discovered in % lo que decía de la seguridad.
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript, however,
\cite{flanagan2006javascript} was introduced in 1997 as a
browser-embedded language and has, since then, become a set of standards
\cite{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an EA on
the browser as early as 1998 \cite{jj-ppsn98}.

@@

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researches to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards}
and DropBox~\cite{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since is a first-classe citizen
in most Platforms as a Service (PaaS) products as most support the Node.js
framework \cite{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and use JSON (JavaScript Object Notation) for data
interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

@@

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library NodEO \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time % [Mario] Creo que si se debería
% dedicar una oración para recordar la diferencia y relación entre NodIO y NodEO
% En mi mente los digo igual, NodEO como Rodeo y NodIO como Radio y me costó trabajo ver la diferencia
% Pero igual un lector que llegue a leer sin vicios, si vea claramente
% la relación.
% OK, lo hago más adelante - JJ
when writing code and leverages %FERGU: qué diferencias hay entre NodIO y NodEO??
expertise acquired in the library itself. %FERGU: tienes info de LoC
                                %usadas para la sección experimental y
                                %justificar esta frase?
% Es el mismo código - JJ

@@

@@

On the other hand, measuring performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. But the raw material of
volunteer computing, number of users and the time spent in the
computation in browser-based volunteer computing experiments, have only been studied in a limited way in
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and Gamma distribution
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

@@

% [Pedro] desde el abstract no se había nombrado "NodIO".
% Creo que en este párrafo debería introducirse, principalmente el nombre.
% Añado una frase presentándolo:
% [Mario] Nota: se menciona también en el segundo párrafo de la
% Introducción.
% OK [JJ]
In this sense, in this paper we propose {\sf NodIO}, a cloud or metal based
volunteer computing system, whose architecture
has been developed using JavaScript on both the client and server sides.
All parts of the system are free and available with a free
license from \url{https://github.com/JJ/splash-volunteer}.
Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several remote %remote requests suena raro, propongo algo como: remote procedures o web interfaces
  requests for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. The information
  is of two kinds: {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
    GET}ing a random chromosome from it, and {\em information} related
  to performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}.
  %The information is of two kinds: Problem Based y ¿Cual otra?
  %¿Logging? No se entiende bien. ¿Podría ser problem domain?
% info. Añadido [JJ]
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with random
  population and sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request). Apart from
  these constants, the algorithm can be run in many different ways:
  stopping when solution is found or, as we have done in this paper,
  using WebWorkers, which run in the background even if the webpage or
  browser tab with the experiment is not currently on focus.
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different web technologies, such as JQuery or chart.js have
been used to build the elements of the system.
%FERGU: Mario, explica en el texto de abajo qué es winston en la figura también.
%Vamos, que todo lo que salga en la figura tiene que ir en el texto en alguna parte :)
% [Mario] Ok voy a explicar la figura. Quería saber primero su opinión tal vez se les hacía mucho detalle.
% Creo que esta figura va mejor en la sección donde hablamos de
% webworkers. La anterior era más adecuada como visión general. [JJ]
% [Mario] Ok
\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/system.pdf}
\caption{Description of the proposed system. Clients execute a JS EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one from the server. }
\label{fig:system}
\end{figure}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First, we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO \cite{nodeo2014}, the basic JavaScript library we have used as base to build {\sf NodIO}. This experiment tries to
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal 512. The algorithm was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments
are shown in Figure \ref{fig:baseline}. %Fergu: dejar claro que es solo una isla

These two experiments establish first a baseline result and show that
it depends on the population. The volunteer computing experiments that we
%FERGU: explicar un poco más lo de "depends on the population"?, por ejemplo añadiendo
%as this is a deceptive problem, and low levels or diversity may not help to
%find the solution \cite{} o algo parecido.
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which this %Ya se defnió antes PaaS
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. No special care was taken to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is, in fact negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPS and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The %FERGU: No debería ser mejor al revés? Cada experimento está compuesto de varias runs. De todas maneras arriba lo has definido así, así que si no quieres no lo cambies. Pero me parece más correcto que un experimento tenga runs, en vez de al revés
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}. This table shows the median number of IPs
intervening in each experiment,  median of the time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
found previously \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same ballpark and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the last two columns, that measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
fair is fair, and we were looking for creating a framework that
allowed to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results is that although it will
happen, one out of roughly 5 or roughly 20, depending on what baseline %FERGU: one->1?
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecessary cap on the time, or number of experiments, that a user %Fergu: user no tiene "an", me lo dijo paloma. es "yuser", no "iuser", por eso no lo tiene :P
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots. From left to right, experiments 4/4, 4/24 and 7/31. Shades
of blue indicate how many experiments included that many unique IPs.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experiment data a bit further to find out why
this happens and also if there is some pattern in the three sets of
experiments. An interesting question to ask, for instance, is if
adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, {\em adding} more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more so taking
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
with a bigger size to check this hypothesis; however, at this point we
have not found this convenient since there are several other issues
that will be solved. %FERGU: and that will be explained next? Algún ejemplo?

@@

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. %Fergu: esto mola, métele más caña en otras secciones para que se vea que es una conclusión util

@@

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even it the user brings forward another tabbed
window.
\item Several instances of NodEO can be executed in parallel in a single web
page. %FERGU: NodEO o NodIO? (es NodEO la instancia, creo pero por si acaso :P)
\item Many implementations of the Web Worker API can use multiple CPUs for
the execution of worker scripts, in particular the execution in multiple CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independently.
\end{itemize}

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP Methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread in locally in the client, these are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implemented in the nodejs %FERGU: creo que oficialmente es Node.js, homogeneizar en el paper
framework. Web request handlers are implemented using the Express web
framework. For keeping a log of the execution the Winston library is used, and
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framework. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the XHR object. %FERGU: Qué es el objeto XHR? Alguna referencia/justificación de por qué se usa?
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}

@@

@@

% The conclusion goes here.

In this paper a client-server architecture for distributed evolutionary algorithms, termed {\sf NodIO}, for volunteer computing has been presented.

In order to establish a baseline performance, the evolutionary algorithm was run in a desktop client-program written in JavaScript using NodEO to solve the 40-trap function.
In this first experiment, it could be observed how many users donated a few cycles and just a few donated a high amount of CPU to the experiment.

As a second experiment, and trying to improve the number of cycles per user, the system architecture was modified. In this case, Web Workers were used so that several clients per browers could be executed.
As a result, in this case a higher number of contributions per volunteer could be observed, although as in the previous experiments a pattern of a few users with many contributions and many with few contributions was obtained.

>>>>

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense than an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript, however,
\cite{flanagan2006javascript} was introduced in 1997 as a
browser-embedded language and has, since then, become a set of standards
\cite{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an EA on
the browser as early as 1998 \cite{jj-ppsn98}.

@@

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researches to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards}
and DropBox~\cite{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since is a first-classe citizen
in most Platforms as a Service (PaaS) products as most support the {\tt node.js}
framework \cite{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and use JSON (JavaScript Object Notation) for data
interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

@@

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library NodEO \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
expertise acquired in the library itself.

@@

On the other hand, measuring performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. But the raw material of
volunteer computing, number of users and the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

@@

@@

In this sense, in this paper we propose {\sf NodIO}, a cloud or metal
based 
volunteer evolutionary computing system based on the {\sf NodEO}
library, whose architecture 
has been developed using JavaScript on the client as well as the
server sides. 
All parts of the system are free and available with a free
license from \url{https://github.com/JJ/splash-volunteer}.
Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} %remote requests suena raro, propongo algo como: remote
          %procedures o web interfaces [jj] changed to routes, its
          %proper name.
  that can be called for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. The information
  is of two kinds: {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
    GET}ing a random chromosome from it, and {\em information} related
  to performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with random
  population and sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request). Apart from
  these constants, the algorithm can be run in many different ways:
  stopping when solution is found or, as we have done in this paper,
  using WebWorkers, which run in the background even if the webpage or
  browser tab with the experiment is not currently on focus.
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different web technologies, such as JQuery or {\tt
  chart.js} have
been used to build the elements of the system.
\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/system.pdf}
\caption{Description of the proposed system. Clients execute a JS EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one from the server. }
\label{fig:system}
\end{figure}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First, we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO
\cite{nodeo2014}, the basic JavaScript library we have used as base to
build {\sf NodIO}, the framework. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal to 512. The algorithm in a
single island was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments
are shown in Figure \ref{fig:baseline}. %Fergu: dejar claro que es solo una isla

These two experiments establish first a baseline result and show that
time to success depends on the population, with bigger population
contributing more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we 
%FERGU: explicar un poco más lo de "depends on the population"?, por ejemplo añadiendo
%as this is a deceptive problem, and low levels or diversity may not help to
%find the solution \cite{} o algo parecido.
%hecho [JJ]
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

@@

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which this %Ya se defnió antes PaaS
experiment could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the months of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. No special care was taken to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is, in fact negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPS and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
%FERGU: No debería ser mejor al revés? Cada experimento está compuesto
%de varias runs. De todas maneras arriba lo has definido así, así que
%si no quieres no lo cambies. Pero me parece más correcto que un
%experimento tenga runs, en vez de al revés [jj] Pues no lo sé. Habría
%que cambiarlo todo. Yo creo que con que sea consisente...
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}. This table shows the median number of IPs
intervening in each experiment,  median of the time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
found previously \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same ballpark and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the last two columns, that measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
fair is fair, and we were looking for creating a framework that
allowed to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results is that although it will
happen, 1 out of roughly 5 or roughly 20, depending on what baseline
%FERGU: one->1? OK [JJ]
experiment you choose, it will not happen always or even a majority of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found; this is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. However, it puts a probably
unnecessary cap on the time, or number of experiments, that a user
%Fergu: user no tiene "an", me lo dijo paloma. es "yuser", no "iuser",
%por eso no lo tiene :P 
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots. From left to right, experiments 4/4, 4/24 and 7/31. Shades
of blue indicate how many experiments included that many unique IPs.}
\label{fig:duration}
\end{figure}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{img/time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
\end{figure}
%
We will have to analyze experiment data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, {\em adding} more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more so taking
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
with a bigger size to check this hypothesis; however, at this point we
have not found this convenient since there are several other issues
that have to be solved, as it will be shown next. %FERGU: and that
                                %will be explained next? Algún
                                %ejemplo? 
% A ver si así - JJ

@@

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. %Fergu: esto mola, métele más caña en otras secciones
                %para que se vea que es una conclusión util 
% Está en las conclusiones - JJ

@@

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even it the user brings forward another tabbed
window.
\item Several evolutionary algorithms can be executed in parallel in a single web
page. %FERGU: NodEO o NodIO? (es NodEO la instancia, creo pero por si
      %acaso :P)
% Mejor evolutionary algorithms - JJ
\item Many implementations of the Web Worker API can use multiple CPUs for
the execution of worker scripts, in particular the execution in multiple CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independently.
\end{itemize}

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this system there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP Methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread in locally in the client, these are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process which is basically
  the same as used previously, except for some minor log-related
  changes. 
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framework. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the XHR object. %FERGU: Qué es el objeto XHR? Alguna referencia/justificación de por qué se usa?
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}

@@

In general, this second version of the {\sf NodIO} framework and the
single experiment performed prove that it can be the foundation for a
high-server-performance distributed evolutionary computation platform,
providing reasonable algorithmic performance and being, in general,
easy to use with a straightforward modification of the fitness
function.

@@

In this paper a client-server architecture for volunteer and distributed
evolutionary algorithms that uses the browser, termed {\sf NodIO} and
based on the {\sf NodEO} evolutionary algorithm library has been evaluated.

In order to establish a baseline performance, the evolutionary
algorithm was run in a desktop client-program written in JavaScript
using NodEO to solve the 40-trap function. The first experiments with
{\sf NodIO} proved that, although obtaining better performance than de
baseline was possible, it did not happen, mainly because of the
difficulty in carrying over volunteers to other experiments when the
one they were participating on was finished, shown in the low
correlation between the number of IPs in successive experiments. This
also resulted in a low number of generations allotted by users.
